Best Practices for Alchemical Free Energy Calculations: v\versionnumber

### 2020-01-19 paper/manuscript.tex
fixes to sec 1 2 and 3

Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor or the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}. 

@@

Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with conventional computational methods, but are more readily tractable with alchemical methods. 
\todo[inline, color={yellow!20}]{JDC: Not sure what is meant by "conventional computational methods". Perhaps we can be precise and say "unbiased molecular dynamics simulations"?}

@@

%such as the entire removal of a ligand from a binding site (and \emph{absolute AFE calculation}), as illustrated in Fig.~\ref{fig:fig_what_is_alchemy} (E). 
alchemical calculations can be used in a variety of scenarios, such as computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (A)); 
\todo[inline, color={yellow!20}]{JDC: Do you intend these to be representative of BLUES type simulations?}
computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (B))~\cite{rustenburg2016measuring, bosisio2016blinded} or compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (C))~\cite{}. 
alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (D))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (E))~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.

@@

Early work used simple (but suboptimal) estimators for this: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}. 
More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{mbar} or removing the need for global equilibrium~\cite{tram}

@@

As the field of molecular simulation moves towards millisecond timescales, accurate alchemical calculations on even more challenging problems will become reasonable to perform. In the meantime, today's users may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change. This best practice guide aims to serve as a go-to reference for current recommendations and tips for users of all experience.  

@@

    \caption{\textbf{Illustration of common types of free energies differences that can be calculated using alchemical free energy methods.} A: Change in free energy due to a conformational change of the molecule. B: Partition coefficient such as $\log P$ or $\log D$ can be computed using alchemical free energy methods. Here, as an example the partition coefficient between methanol and water is shown. C: Free energy difference associated with the insertion of a molecule into a membrane. D: Effect of mutations of protein or host residues on free energies of binding.  E: Absolute free energies of binding of a small molecule to a host (e.g. protein), F: Relative free energy of binding of one molecule with respect to another, here toluene and benzyl alcohol.

@@

In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:step1}), choice and use of correct protocols~(\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used for the analysis of alchemical calculations~(\ref{sec:data_analysis}). 

@@

    \textit{CPU} - Central Processing Unit\\
    \textit{BAR} - Bennett Acceptance Ratio\\
    \textit{FEP} - Free Energy Perturbation\\
    \textit{GPCR} - G-Protein Coupled Receptor\\
    \textit{GPU} - Graphics Processing Unit\\
    \textit{MBAR} - Multistate Bennett Acceptance Ratio\\
    \textit{MCSS} - Maximum Common Substructure\\
    \textit{MD} - Molecular Dynamics\\
    \textit{RBFE} - Relative Binding Free Energy\\
    \textit{RMSE} - Root Mean Square Error\\
    \textit{SAR} - Structure-Activity Relationships\\
    \textit{TI} - Thermodynamic Integration

@@

U - Potential energy\\

@@

\begin{tcolorbox}[title=Common terminology, colback=green!10!white]
    ligand - small molecular binding to a larger host molecule such as a protein.\\ 

@@

% Step 1   %

@@

\label{sec:step1}

@@

Generally, binding modes within congeneric series are conserved~\cite{}, however, exceptions exist~\cite{}, and weaker binding compounds are more likely to adopt different binding modes (REFS). There may also be ambiguity about the relative orientation of substituents. A known issue involves a 180 degree flip in the dihedral angle of a 6-membered aromatic ring leading to a different spatial position of ortho- or meta- substituents that otherwise should overlap within a series. The 180 degree flip of the ring may not occur enough during simulations to overcome the starting configuration bias. Another scenario may be equatorial and axial substituted saturated rings (e.g. cyclohexane derivatives). This situation may be addressed by explicitly modelling different binding modes of the same ligand and combining later computed free energy differences for different binding modes into a relative free energies of binding~\cite{kaus2015how}.

@@

Input coordinates for a congeneric series may be generated by docking calculations, or by ligand alignment using maximum common substructure algorithms. The latter tends to produce alignments that are more conserved and more consistent free energy changes across a dataset, but will struggle to yield reasonable results for relative binding free energy calculations that involve a significant binding mode rearrangement. This may also lead to steric clashes with the receptor coordinates of the reference ligand if structural rearrangements are needed to accommodate different members of the congeneric series. Small steric clashes may be resolved during subsequent simulation equilibration prior to data collection but there is a risk that the complex relaxes to an alternative metastable state. 

@@

Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straight forward. For example, a 3D minimised structure of the solute can be generated with a simple tool such as babel and solvated to prepare the input  to compute a free energy of hydration. However, in these cases a careful choice of forcefield, as well as water models or organic solvents is essential. See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion on these choices. 

@@

Several other case studies of FEP calculations can be found in the review by Williams-Noonan et al.~\cite{williams-noonan2018free}. For comparison of FEP+ and Gromacs (using the AMBER99SB-ILDN and GAFF2 force field), cf. the recently published study by Pérez-Benito et al.~\cite{perez-benito2019predicting}.
An overview of further suggested benchmark sets can be found in the review by Mobley and Gilson~\cite{mobley2017predicting} or on \url{alchemistry.org}~\cite{alchemistry}. These include cyclodextrins, the Cytochrome C peroxidase (CCP) protein model binding site, thrombin and bromodomains as well as solvation benchmark sets~\cite{paliwal2011benchmark}. Further work on shared protein-ligand benchmark sets is ongoing at \url{https://github.com/openforcefield/PLBenchmarks}.

@@

% \begin{checklist}{Simulation preparation}
% \textbf{How do I get started setting up an alchemical free energy calculation}
% Extensive explanation for the checklist questions can be found in Section~\ref{sec:step1}.
% \begin{itemize}
% \item Have I followed the Best practices for biomolecular simulation set up?
% \item In a relative simulation, will I run into problems with clashing geometries in the ligand transformation or crystal waters?
% \end{itemize}
% \end{checklist}
% \end{Checklists*}

@@

\section*{Author Contributions}
%%%%%%%%%%%%%%%%
% This section mustt describe the actual contributions of
% author. Since this is an electronic-only journal, there is
% no length limit when you describe the authors' contributions,
% so we recommend describing what they actually did rather than
% simply categorizing them in a small number of
% predefined roles as might be done in other journals.
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io
% for more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%%
\todo[inline, color={green!20}]{ASJSM: @everyone, can you check your contribution is acknowledge and if not add the appropriate information. }
(Explain the contributions of the different authors here)
ASJSM: Coordinated the document, contributed to most sections, and co-designed figures ~\ref{} and created figures \ref{fig:fig_what_is_lambda},~\ref{fig:fig1_what_is_alchemy},~\ref{fig:overlap} and replotted~\ref{fig:automatic-equilibration-detection} and improved ~\ref{fig:fig_types_of_networks}.
JS: created figures \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, the data representation section, the checklist and contributed to general formatting discussions and editing.
MK: Contributed to the Data Analysis section, provided the data for figure~\ref{fig:overlap} and helped edit the paper.
DLM: Contributed to the outline, drafted some of the sections, gave ideas on figures, and helped edit the paper.
GT: Contributed to introduction and drug discovery sections, and helped edit the paper.
AR: Created figure~\ref{fig:freeenergytrajectories}, contributed to section \ref{sec:simulation_protocol_choice}, and helped edit the paper.
MRS: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.
LNN: Helped write the simulation length, stopping conditions, and information saving section.
BA: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.
JM: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:step1},~\ref{} 4.2, 5.3, 6, 8.3.1, 8.3.2\
JDC: Wrote section~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \todo[inline, color={red!20}]{@JDC check info}
HX: Contributed section~\ref{subsec:accuracy}, to section~\ref{sec:relative-fe-protocol}, and to section~\ref{sec:are-they-good}.

>>>>

Alchemical free energy calculations compute free energy differences associated with transfer processes, such as the binding of a small molecule to a receptor, the transfer of a small molecule from an aqueous to apolar phase~\cite{zwanzig1954hightemperature}, or effects of protein side chain mutations on e.g. binding affinities.. 

@@

Fig.~\ref{fig:fig_what_is_alchemy} illustrates common free energy changes that may be difficult to compute with unbiased molecular dynamics methods, but are more readily tractable with alchemical methods.

@@

Alchemical calculations can be used in a variety of scenarios, such as computing the free energy of a conformational change for a molecule with a high barrier to interconversion (Fig.~\ref{fig:fig_what_is_alchemy} (A)); 
\todo[inline, color={yellow!20}]{JDC: Do you intend these to be representative of BLUES type simulations? -- yes}
computing partition ($\log P$) or distribution ($\log D$) coefficients between environments (Fig.~\ref{fig:fig_what_is_alchemy} (B))~\cite{rustenburg2016measuring, bosisio2016blinded} or compartments or membranes (Fig.~\ref{fig:fig_what_is_alchemy} (C))~\cite{corey2019insights}. 
Furthermore, alchemical calculations are frequently used to estimate changes in free energies upon modifying a ligand or protein: a protein residue can be alchemically mutated to probe the impact on binding affinity (Fig.~\ref{fig:fig_what_is_alchemy} (D))\cite{hauser2018predicting,aldeghi2018accurate} or changes in protein thermostability~\cite{seeliger2010protein,gapsys2016insights,gapsys2016accurate,aldeghi2019accurate}; the entire ligand can be alchemical transferred from protein to solvent in an absolute binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (E))~\cite{mobley2007predicting,aldeghi2015accurate,aldeghi2017predictions}; and small alchemical modifications can be made between chemically related ligands to estimate differences in binding free energies (Fig.~\ref{fig:fig_what_is_alchemy} (F))~\cite{wang2015accurate,mey2016blinded,gapsys2020large}.

@@

Early work used simple (but suboptimal) estimators for this: free energy perturbation (FEP) using the Zwanzig relation~\cite{zwanzig1954hightemperature} or thermodynamic integration (TI), for which the theory dates back decades but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}. %
More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}

@@

As the field of molecular simulation moves towards millisecond timescales, accurate alchemical calculations on even more challenging problems will become reasonable to perform. In the meantime, today's users may find it hard to get started with these complex calculations whilst also keeping up with the fast pace of change. This Best Practice guide aims to serve as a go-to reference for current recommendations and tips for users of all experience.  

@@

    \caption{\textbf{Illustration of common types of free energies differences that can be calculated using alchemical free energy methods.} \textbf{A}: Change in free energy due to a conformational change of the molecule. \textbf{B}: Partition coefficient such as $\log P$ or $\log D$ depend on a change in free energy between different phases; here, as an example the partition coefficient between methanol and water is shown. \textbf{C}: Free energy difference associated with the insertion of a molecule into a membrane. \textbf{D]: Effect of mutations of protein or host residues on free energies of binding. \textbf{E}: Absolute free energies of binding of a small molecule to a host (e.g. protein), \textbf{F}: Relative free energy of binding of one molecule with respect to another, here toluene and benzyl alcohol.

@@

In the remainder of this paper, we will cover topics that are key to the preparation~(\ref{sec:prerequisites}), choice and use of correct protocols~(\ref{sec:simulation_protocol_choice}), and finally the best practices that should be used for the analysis of alchemical calculations~(\ref{sec:data_analysis}). 

@@

    \textit{CPU} -- Central Processing Unit\\
    \textit{BAR} -- Bennett Acceptance Ratio\\
    \textit{FEP} -- Free Energy Perturbation\\
    \textit{GPCR} -- G-Protein Coupled Receptor\\
    \textit{GPU} -- Graphics Processing Unit\\
    \textit{MBAR} -- Multistate Bennett Acceptance Ratio\\
    \textit{MCSS} -- Maximum Common Substructure\\
    \textit{MD} -- Molecular Dynamics\\
    \textit{RMSE} -- Root Mean Square Error\\
    \textit{MUE} -- Mean unsigned error\\
    \textit{SAR} -- Structure-Activity Relationships\\
    \textit{TI} -- Thermodynamic Integration

@@

$U$ -- potential energy\\
$u$ -- reduced potential energy \\
$\Delta G$ -- free energy \\
$\Delta f$-- reduced free energy \\
$k_B$-- Boltzmann constant \\
$T$ -- temperature
$k_b$ -- rate of binding \\
$K_D$ -- dissociation constant \\
$\lambda$ -- parameter for regulation of the alchemcial free energy path \\
$g$ -- statistical inefficiency

@@

\begin{tcolorbox}[title=Common terminology, colback=yellow!10!white]
    ligand - small molecular binding to a larger host molecule such as a protein (recpetor).\\ 

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Simulation prerequisites  %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\label{sec:prerequisites}

@@

Generally, binding modes within congeneric series are conserved~\cite{wacker2010conserved}, however, exceptions exist~\cite{brandt2011congeneric,nazare2005probing}, and weaker binding compounds are more likely to adopt different binding modes (REFS). There may also be ambiguity about the relative orientation of substituents. A known issue involves a 180 degree flip in the dihedral angle of a 6-membered aromatic ring leading to a different spatial position of ortho- or meta- substituents that otherwise should overlap within a series. The 180 degree flip of the ring may not occur enough during simulations to overcome the starting configuration bias. Another scenario may be equatorial and axial substituted saturated rings (e.g. cyclohexane derivatives). This situation may be addressed by explicitly modelling different binding modes of the same ligand and combining later computed free energy differences for different binding modes into a relative free energies of binding~\cite{kaus2015how}.

@@

Input coordinates for a congeneric series may be generated by docking calculations, or by ligand alignment using maximum common substructure algorithms. The latter tends to produce alignments that are more conserved and more consistent free energy changes across a dataset, but will struggle to yield reasonable results for relative binding free energy calculations that involve a significant binding mode rearrangement. This may also lead to steric clashes with the receptor coordinates of the reference ligand if structural rearrangements are needed to accommodate different members of the congeneric series. Small steric clashes may be resolved during subsequent simulation equilibration prior to data collection, but there is a risk that the complex relaxes to an alternative metastable state. 

@@

Preliminary considerations necessary for using free energy methods to compute partition coefficients are generally more straight forward. For example, a 3D minimised structure of the solute can be generated with a simple tool such as openBabel and solvated to prepare the input  to compute a free energy of hydration~\cite{oboyle2011open}. However, in these cases a careful choice of forcefield, as well as water models or organic solvents is essential. See for example~\cite{bosisio2016blinded,rustenburg2016measuring} for a good discussion on these choices. 

@@

Several other case studies of alchemical calculations can be found in the review by Williams-Noonan et al.~\cite{williams-noonan2018free}. For comparison of FEP+ and Gromacs (using the AMBER99SB-ILDN and GAFF2 force field), cf. the recently published study by Pérez-Benito et al.~\cite{perez-benito2019predicting}.
An overview of further suggested benchmark sets can be found in the review by Mobley and Gilson~\cite{mobley2017predicting} or on \url{alchemistry.org}~\cite{alchemistry}. These include cyclodextrins, the Cytochrome C peroxidase (CCP) protein model binding site, thrombin and bromodomains as well as solvation benchmark sets~\cite{paliwal2011benchmark}. Further work on shared protein-ligand benchmark sets is ongoing at \url{https://github.com/openforcefield/PLBenchmarks}.

@@

% \begin{checklist}{Simulation preparation}
% \textbf{How do I get started setting up an alchemical free energy calculation}
% Extensive explanation for the checklist questions can be found in Section~\ref{sec:prerequisites}.
% \begin{itemize}
% \item Have I followed the Best practices for biomolecular simulation set up?
% \item In a relative simulation, will I run into problems with clashing geometries in the ligand transformation or crystal waters?
% \end{itemize}
% \end{checklist}
% \end{Checklists*}

@@

\section*{Author Contributions}
%%%%%%%%%%%%%%%%
% This section mustt describe the actual contributions of
% author. Since this is an electronic-only journal, there is
% no length limit when you describe the authors' contributions,
% so we recommend describing what they actually did rather than
% simply categorizing them in a small number of
% predefined roles as might be done in other journals.
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io
% for more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%%
\todo[inline, color={green!20}]{ASJSM: @everyone, can you check your contribution is acknowledge and if not add the appropriate information. }
(Explain the contributions of the different authors here)
ASJSM: Coordinated the document, contributed to most sections, and co-designed figures ~\ref{} and created figures \ref{fig:fig_what_is_lambda},~\ref{fig:fig1_what_is_alchemy},~\ref{fig:overlap} and replotted~\ref{fig:automatic-equilibration-detection} and improved ~\ref{fig:fig_types_of_networks}.
JS: created figures \ref{fig:fig_what_is_alchemy},~\ref{fig:fig_binding_thermodynamic_cycle},~\ref{fig:fig_topology},~\ref{fig:fig_mcss},~\ref{fig:fig_types_of_networks},~\ref{fig:fig_absolute_thermodynamic_cycle},~\ref{fig:scatterplot_analysis}, the data representation section, the checklist and contributed to general formatting discussions and editing.
MK: Contributed to the Data Analysis section, provided the data for figure~\ref{fig:overlap} and helped edit the paper.
DLM: Contributed to the outline, drafted some of the sections, gave ideas on figures, and helped edit the paper.
GT: Contributed to introduction and drug discovery sections, and helped edit the paper.
AR: Created figure~\ref{fig:freeenergytrajectories}, contributed to section \ref{sec:simulation_protocol_choice}, and helped edit the paper.
MRS: Helped create figure~\ref{fig:fig_what_is_lambda}, wrote sections describing choices for alchemical pathways and analysis for free energy calculations.
LNN: Helped write the simulation length, stopping conditions, and information saving section.
BA: Helped write the uncertainty estimation, stopping conditions, and output analysis sections and created figure ~\ref{fig:convergence_forward_reverse}.
JM: Contributed to sections~\ref{subsec:reproducible},~\ref{sec:prerequisites},~\ref{} 4.2, 5.3, 6, 8.3.1, 8.3.2\
JDC: Wrote section~\ref{sec:decorrelating-samples} and~\ref{sec:automatic-equilibration-detection} discussed structure and design of the whole document, suggested Figs.~\ref{fig:fig_what_is_alchemy} and~\ref{fig:fig_sampling_scheme} \todo[inline, color={red!20}]{@JDC check info}
HX: Contributed section~\ref{subsec:accuracy}, to section~\ref{sec:relative-fe-protocol}, and to section~\ref{sec:are-they-good}.
